Equalizer Programming Guide

### 2009-03-23 Developer/ProgrammingGuide/paper.tex
PG: Document layout API
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

\date{
  \vspace{15cm}
  {\Large\sffamily The official reference for creating parallel,
    scalable OpenGL\texttrademark\ applications with the Equalizer
    parallel rendering framework}\\\vspace{1cm}
  Version 1.5.5 for Equalizer 0.6+svn\\[\medskipamount]
%  \today
}

@@

\clearpage
\pagenumbering{roman}
\pagestyle{headings}
\tableofcontents
\listoffigures
\vfill{\center\begin{tabularx}{\textwidth}{|l|l|X|}
    \hline
    \bf Rev & \bf Date     & \bf Changes \\
    \hline
    1.0     & Oct 28, 2007 & Initial Version for Equalizer 0.4\\
    1.2     & Apr 15, 2008 & Revision for Equalizer 0.5\\
    1.4     & Nov 25, 2008 & Revision for Equalizer 0.6\\
    1.5.1   & Dec 08, 2008 & Minor updates based on the print version\\
    1.5.2   & Jan 10, 2009 & Added \sref{sCompositor}\\
    1.5.3   & Jan 13, 2009 & Added \sref{sNetwork}\\
    1.5.4   & Feb 09, 2009 & Added \sref{sNetUsage}\\
    1.5.5   & Mar 10, 2009 & Updated to reflect new AGL event dispatch\\
    \hline
  \end{tabularx}}
\clearpage

@@

\subsection{Canvases}

@@

\subsection{Layouts}

@@

@@

\begin{figure}[ht!]\center
  \includegraphics[width=\textwidth]{images/uml}
  {\caption{\label{fUml}UML diagram eqPly and relevant Equalizer classes.}}
\end{figure}

@@

{\footnotesize\begin{lstlisting}
class NodeFactory : public eq::NodeFactory
{
public:
    virtual eq::Config*  createConfig( eq::ServerPtr parent )
        { return new eqPly::Config( parent ); }
    virtual eq::Node*    createNode( eq::Config* parent )  
        { return new eqPly::Node( parent ); }
    virtual eq::Pipe*    createPipe( eq::Node* parent )
        { return new eqPly::Pipe( parent ); }
    virtual eq::Window*  createWindow( eq::Pipe* parent )
        { return new eqPly::Window( parent ); }
    virtual eq::Channel* createChannel( eq::Window* parent )
        { return new eqPly::Channel( parent ); }
};
\end{lstlisting}}

@@

Distributed objects are created by subclassing from
\textsf{eq::net::Object}. Distributed objects can be static (immutable) or
dynamic. Dynamic objects are versioned.

@@

Equalizer determines the way changes are to be handled by calling
\textsf{Object::getChangeType} during the registration of the master
version of a distributed object. The change type determines the memory
footprint and the contract for calling the serialization methods. The
following change types are possible:

@@

Versioned objects have to override \textsf{getChangeType} to indicate
how they want to have changes to be handled. All versioned objects
currently implemented have the following characteristics:
\begin{itemize}
\item Only the master instance of the object is writable, that is,
  \textsf{eq::net::Object::com\-mit} can be called only on the master
  instance to generate a new version.
\item Slave instance versions can only be advanced, that is,
  \textsf{eq::net::Object::sync( version )} with a version smaller than
  the current version will fail.
\item Newly mapped slave instance are mapped to the oldest available
  version by default, or to the specified version in
  \textsf{Session::mapObject}.
\end{itemize}

@@

\textsf{eqPly} does not track changes in the \textsf{FrameData} since it
is relatively small and changes frequently. Therefore, the instance and
delta data are the same, and \textsf{pack} and \textsf{unpack} are not
implemented. The change type is \textsf{INSTANCE}. The serialization
functions exploit the fact that the frame data is layed out in one
anonymous structure in memory, and transmit all the data in one call:

{\footnotesize\begin{lstlisting}
void FrameData::getInstanceData( eq::net::DataOStream& os )
{ 
    os.writeOnce( &data, sizeof( data )); 
}

void FrameData::applyInstanceData( eq::net::DataIStream& is )
{
    memcpy( &data, is.getRemainingBuffer(), sizeof( data ));
    is.advanceBuffer( sizeof( data ));
}
\end{lstlisting}}%>>

@@

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/modelDist.pdf}
  {\caption{\label{fModelDist}Scene Data in eqPly}}
\end{wrapfigure}
\textsf{eqPly} uses static distributed objects to distribute the model
loaded by the application. This model can be easily extended to
versioned objects to support dynamic data changes.

The kD-tree data structure and rendering code for the model is strongly
separated from Equalizer, and kept in the separate namespace
\textsf{mesh}. It could also be used in other rendering software, for
example in a GLUT application. To keep this separation, an external
'mirror' hierarchy is constructed around the tree. This hierarchy of
\textsf{VertexBufferDist} nodes is responsible for cloning the model
data on the remote render clients.

The identifier of the root object of this distributed hierarchy is passed
as part of the \textsf{InitData}, and used during remote node
initialization to map the model. \fig{fModelDist} shows the UML
hierarchy of the model and distribution classes.

Each \textsf{VertexBufferDist} object has one node of the model's data
tree. It is serializing the data for this node. Furthermore, it mirrors
the kD-tree by having a \textsf{VertexBufferDist} child for each child
of its corresponding tree node. During serialization, the identifier of
these children is sent to the remote nodes, which reconstruct the mirror
distribution hierarchy and model data tree based on this data.

@@

\begin{wrapfigure}{r}{.618\textwidth}
%\begin{figure}[ht!]\center
  \includegraphics[width=.618\textwidth]{images/objects.pdf}
  {\caption{\label{fObjects}Scene Graph Distribution}}
\end{wrapfigure}
Applications distributing a dynamic scene graph use the frame data
instead of the init data as the entry point to their scene graph data
structure. \fig{fObjects} shows one possible implementation, where the
identifier and version of the scene graph root are transported using the
frame data. The scene graph root then serializes and de-serializes his
immediate children by transferring their identifier and current version,
similar to the static distribution done by \textsf{eqPly}.

@@

\begin{figure}[ht!]\center
%\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.9\textwidth]{images/configInit.pdf}
  {\caption{\label{fConfigInit}Config Initialization Sequence}}
\end{figure}
The config initialization happens in parallel, that is, all config
initialization tasks are transmitted by the server at once and their
completion is synchronized afterwards.

The tasks are executed by the node and pipe threads in parallel. The
parent's initialization methods are always executed before any child
initialization method. This parallelization is necessary to allow a
speedy startup of the configuration on large-scale graphics clusters. On
the other hand, it means that initialization functions are called even
if the parent's initialization has failed. \fig{fConfigInit} shows a
sequence diagram of the config initialization.

@@

@@

For each active render client, one \textsf{eq::Node} instance is
created on the appropriate machine. Nodes are only instantiated on their
render client processes, i.e., each process should only have one
instance of the \textsf{eq::Node} class. The application process might
also have a node class, which is handled in exactly the same way as the
render client nodes. The application and render clients might use a
different node factory, instantiating a different type of \textsf{eq::Node}.

@@

The actual mapping of the static data is done by the config. The config
first retrieves the distributed \textsf{InitData}. The object is
directly unmapped since it is static, and the data has been retrieved.

The init data contains, among other things, the identifier of the
root of the model's kD-tree. Using this identifier, the model is mapped
to the node, as explained in \sref{sSceneData}:

{\footnotesize\begin{lstlisting}
bool Config::mapData( const uint32_t initDataID )
{
    if( _initData.getID() == EQ_ID_INVALID )
    {
        const bool mapped = mapObject( &_initData, initDataID );
        EQASSERT( mapped );
        unmapObject( &_initData ); // data was retrieved, unmap immediately
    }
    else  // appNode, _initData is registered already
        EQASSERT( _initData.getID() == initDataID );

    const uint32_t modelID = _initData.getModelID();
    if( modelID == EQ_ID_INVALID ) // no model loaded by application
        return true;

    if( !_model )
        _model = ModelDist::mapModel( this, modelID );

    return (_model != 0);
}
\end{lstlisting}}

Since the model data is static, the mirror tree for data distribution
is only created temporarily to retrieve the data. This is done in a
static function of the \textsf{VertexBufferDist} class. One instance of
the distribution class is mapped to the model's identifier. During
application of the instance data, it will map its children recursively:

{\footnotesize\begin{lstlisting}
mesh::VertexBufferRoot* VertexBufferDist::mapModel(eq::net::Session* session,
                                                   const uint32_t modelID )
{
    VertexBufferDist modelDist( 0, 0 );

    if( !session->mapObject( &modelDist, modelID ))
    {
        EQWARN << "Mapping of model failed" << endl;
        return 0;
    }

    modelDist._unmapTree(); // No longer needed

    return const_cast< mesh::VertexBufferRoot* >( modelDist._root );
}
\end{lstlisting}}%>>

@@

The object manager is not strictly a part of the window. It
is mentioned here since the \textsf{eqPly} window uses an object manager.

@@

\textsf{eqPly} uses the window's object manager in the rendering code to
obtain the OpenGL objects for a given data item. The address of the data
item to be rendered is used as the key. All objects managed by the OM
are reference counted. If an application releases the objects properly,
they are automatically de-allocated. It is also possible to manually
manage de-allocation of objects, which might be more convenient in some
cases.

Currently, support for display lists, VBO's, textures and shaders is
implemented. For each object, the following functions are available:

\begin{description}
\item[supportsObjects()] returns true if the usage for this particular
  type of objects is supported. For objects available in OpenGL 1.1 or
  earlier, this function is not implemented.
\item[getObject( key )] returns the object associated with the given
  key, or FAILED. Increases the reference count of existing objects.
\item[newObject( key )] allocates a new object for the given
  key. Returns FAILED if the object already exists or if the allocation
  failed. Sets the reference count of a newly created object to one.
\item[obtainObject( key )] convenience function which gets or obtains
  the object associated with the given key. Returns FAILED only if the
  object allocation failed.
\item[releaseObject( key \textbar\ name )] decreases the reference count and
  deletes the object if the reference count reaches zero.
\item[deleteObject( key \textbar\ name )] manually deletes the object. To be
  used if reference counting is not used.
\end{description}

@@

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/DB.png}
  {\caption{\label{fDBDest}Destination view of a DB compound}}
\end{wrapfigure}
After the basic view setup, a directional light is configured, and the
model is positioned using the camera parameters from the frame data. The
camera parameters are transported using the the frame data to ensure
that all channels render a given frame using the same position.

@@

    Config*          config = static_cast< Config* >( getConfig( ));
    const Model*     model  = config->getModel();
    const eq::Range& range  = getRange();

@@

The rendering loop maintains a list of candidates to render, which
initially contains the root node. Each candidate of this list is tested
for full visibility against the frustum and range, and rendered if
visible. It is dropped if it is fully invisible or fully out of
range. If it is partially visible or partially in range, the children of
the node are added to the candidate list. \fig{fRender} shows a flow
chart of the rendering algorithm, which performs efficient view frustum
and range culling.

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/render.pdf}
  {\caption{\label{fRender}Main Render Loop}}
\end{wrapfigure}
The actual rendering uses display lists or vertex buffer objects. These
OpenGL objects are allocated using the object manager. The rendering is
done by the leaf nodes, which are small enough to store the vertex
indices in a \textsf{short} value for optimal performance with VBO's.
The leaf nodes reuse the objects stored in the object manager, or create
and set up new objects if it was not yet set up. Since one object
manager is used per thread (pipe), this allows a thread-safe sharing of
the compiled display lists or VBO's across all windows of a pipe.

@@

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/eventFilter.pdf}
  {\caption{\label{fEventProcessing}Event Processing}}
\end{wrapfigure}
The purpose of the \textsf{OSWindow} processing method,
\textsf{processEvent}, is to perform window-system specific event
processing. For example, \textsf{AGLWindow::processEvent} calls
\textsf{aglUpdateContext} whenever a resize event is received. For
further, generic processing, the \textsf{Event} is passed on to
\textsf{Window::processEvent}. This \textsf{Event} no longer contains
the native event.

\textsf{Window::processEvent} is responsible for either handling
the event locally, or for translating it into a generic
\textsf{ConfigEvent}. The config events are send to the application
thread using \textsf{Config::sendEvent}.

@@

The concrete implementation of these window interfaces is responsible
for setting up and de-initializing event handling.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/eventUML.pdf}
  {\caption{\label{fEventUML}UML Diagram for Event Handling Classes}}
\end{wrapfigure}
The glX event handler iterates over all windows of the registered pipe
to find the originating \textsf{GLXWindowIF}, which is then used to
process the event in a fashion similar to AGL. The
\textsf{GLXWindowEvent} passes the \textsf{XEvent} to
\textsf{GLXWindowIF::\-process\-Event}.

@@

During this initialization, Equalizer creates and manages two threads
for communication, the receiver thread and the command thread. Normally
no application code is executed from these two threads.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/threads.pdf}
  {\caption{\label{fThreads}Threads within one node process}}
\end{wrapfigure}
The receiver and command thread are terminated when the application
stops network communications using \textsf{Client::exitLocal}.

@@

The application has extended control over the task synchronization
during a frame. Upon \textsf{Config::startFrame}, Equalizer invokes the
\textsf{frameStart} task methods of the various entities. The entities
unlock all their children by calling \textsf{startFrame}, e.g.,
\textsf{Node::frameStart} has to call \textsf{Node::startFrame} in order
to unlock the pipe threads. Note that certain \textsf{startFrame} calls,
e.g., \textsf{Window::startFrame}, are currently empty since the
synchronization is implicit due to the sequential execution within the
thread.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/frameSync.pdf}
  {\caption{\label{fFrameSync}Per-Node Frame Synchronization}}
\end{wrapfigure}
The local synchronization consists of \textsf{releaseFrameLocal} to
unlock the local frame, and of \textsf{waitFrameLocal} to wait for the
unlock. For the default synchronization model \textsf{sync\_draw},
Equalizer uses the task method \textsf{frameDrawFinish} which is called
on each resource after the last \textsf{Channel::frameDraw} invocation
for this frame. Consequently, \textsf{Pipe::frameDraw\-Fi\-nish} calls
\textsf{Pipe::\-release\-Frame\-Lo\-cal} to signal that it is done
drawing the current frame, and \textsf{Node::frameDrawFinish} calls
\textsf{Pipe::waitFrameLocal} for each of its pipes to block the node
thread until the current frame has been drawn.

@@

\begin{enumerate}
\item\textsf{getWGLPipeDC} is used to get an affinity device context,
  which might be needed for window creation. If a context is returned, a
  function pointer to delete the returned device context is set as
  well. This function might return 0, which is not an error, but
  signifies that no affinity context is used.
\item\textsf{chooseWGLPixelFormat} chooses a pixel format based on the
  window attributes. If no device context is given, it uses the system
  device context. The chosen pixel format is set on the passed device
  context.
\item\textsf{configInitWGLDrawable} creates the drawable. The device
  context passed to \textsf{configInitWGLDrawable} is used to query the
  pixel format and is used as the device context for creating a
  PBuffer. If no device context is given, the display device context is
  used. On success, it sets the window handle. Setting a window handle
  also sets the window's device context.
\item\textsf{createWGLContext} creates an OpenGL rendering context using
  the given device context. If no device context is given, the window's
  device context is used. This function does not set the window's OpenGL
  context.
\end{enumerate}

@@

    if( !context )
    {
        if( dc )
            deleteDCProc( dc );
        return false;
    }

>>>>

\date{
  \vspace{15cm}
  {\Large\sffamily The official reference for creating parallel,
    scalable OpenGL\texttrademark\ applications with the Equalizer
    parallel rendering framework}\\\vspace{1cm}
  Version 1.5.6 for Equalizer 0.6+svn\\[\medskipamount]
%  \today
}

@@

\clearpage
\pagenumbering{roman}
\pagestyle{headings}
\tableofcontents
\listoffigures
\vfill{\center\begin{tabularx}{\textwidth}{|l|l|X|}
    \hline
    \bf Rev & \bf Date     & \bf Changes \\
    \hline
    1.0     & Oct 28, 2007 & Initial Version for Equalizer 0.4\\
    1.2     & Apr 15, 2008 & Revision for Equalizer 0.5\\
    1.4     & Nov 25, 2008 & Revision for Equalizer 0.6\\
    1.5.1   & Dec 08, 2008 & Minor updates based on the print version\\
    1.5.2   & Jan 10, 2009 & Added \sref{sCompositor}\\
    1.5.3   & Jan 13, 2009 & Added \sref{sNetwork}\\
    1.5.4   & Feb 09, 2009 & Added \sref{sNetUsage}\\
    1.5.5   & Mar 10, 2009 & Updated to reflect new AGL event dispatch\\
    1.5.6   & Mar 23, 2009 & Documented Layout API\\
    \hline
  \end{tabularx}}
\clearpage

@@

\subsection{\label{sCanvas}Canvases}

@@

\subsection{\label{sLayout}Layouts}

@@

\begin{figure}[ht!]\center
  \includegraphics[width=.9\textwidth]{images/uml}
  {\caption{\label{fUml}UML diagram eqPly and relevant Equalizer classes.}}
\end{figure}

@@

@@

{\footnotesize\begin{lstlisting}
class NodeFactory : public eq::NodeFactory
{
public:
    virtual eq::Config*  createConfig( eq::ServerPtr parent )
        { return new eqPly::Config( parent ); }
    virtual eq::Node*    createNode( eq::Config* parent )  
        { return new eqPly::Node( parent ); }
    virtual eq::Pipe*    createPipe( eq::Node* parent )
        { return new eqPly::Pipe( parent ); }
    virtual eq::Window*  createWindow( eq::Pipe* parent )
        { return new eqPly::Window( parent ); }
    virtual eq::Channel* createChannel( eq::Window* parent )
        { return new eqPly::Channel( parent ); }
    virtual eq::View* createView() { return new eqPly::View(); }
};
\end{lstlisting}}

@@

Distributed objects are created by subclassing from \textsf{eq::Object}
or \textsf{eq::net::Object}. Distributed objects can be static
(immutable) or dynamic. Dynamic objects are versioned.

@@

Equalizer determines the way changes are to be handled by calling
\textsf{Object::get\-Change\-Type} during the registration of the master
version of a distributed object. The change type determines the memory
footprint and the contract for calling the serialization methods. The
following change types are possible:

@@

Versioned objects have to override \textsf{getChangeType} to indicate
how they want to have changes to be handled. All versioned objects
currently implemented have the following characteristics:

\begin{itemize}
\item Only the master instance of the object is writable, that is,
  \textsf{eq::net::Object::com\-mit} can be called only on the master
  instance to generate a new version.
\item Slave instance versions can only be advanced, that is,
  \textsf{eq::net::Object::sync( version )} with a version smaller than
  the current version will fail.
\item Newly mapped slave instance are mapped to the oldest available
  version by default, or to the specified version in
  \textsf{Session::mapObject}.
\end{itemize}

@@

The \textsf{eq::Object} implements one possible way of tracking data
changes in versioned objects using the concept of dirty bits, which
facilitates the implementation of the data distribution code.

The \textsf{FrameData} object tracks changes by setting the appropriate
dirty bit whenever some of its data is changed. The serialization
methods are called by the \textsf{eq::Object} with the dirty bit mask
needed to serialize all data or the changes since the last
\textsf{commit}. The \textsf{FrameData} only defines its own dirty bits
and serialization code:

{\footnotesize\begin{lstlisting}
/** The changed parts of the data since the last pack(). */
enum DirtyBits
{
    DIRTY_CAMERA = eq::Object::DIRTY_CUSTOM << 0,
    DIRTY_FLAGS  = eq::Object::DIRTY_CUSTOM << 1,
    DIRTY_VIEW   = eq::Object::DIRTY_CUSTOM << 2,
};

void FrameData::serialize( eq::net::DataOStream& os, const uint64_t dirtyBits )
{
    eq::Object::serialize( os, dirtyBits );
    if( dirtyBits & DIRTY_CAMERA )
        os << _translation << _rotation;
    if( dirtyBits & DIRTY_FLAGS )
        os << _modelID << _renderMode << _color << _ortho << _statistics
           << _help << _wireframe;
    if( dirtyBits & DIRTY_VIEW )
        os << _currentViewID;
}

void FrameData::deserialize( eq::net::DataIStream& is,
                             const uint64_t dirtyBits )
{
    eq::Object::deserialize( is, dirtyBits );
    if( dirtyBits & DIRTY_CAMERA )
        is >> _translation >> _rotation;
    if( dirtyBits & DIRTY_FLAGS )
        is >> _modelID >> _renderMode >> _color >> _ortho >> _statistics
           >> _help >> _wireframe;
    if( dirtyBits & DIRTY_VIEW )
        is >> _currentViewID;
}
\end{lstlisting}}%>>

@@

\textsf{eqPly} uses static distributed objects to distribute the model
loaded by the application. This approach can be easily extended to
versioned objects to support dynamic data modifications.

The kD-tree data structure and rendering code for the model is strongly
separated from Equalizer, and kept in the separate namespace
\textsf{mesh}. It can also be used in other rendering software, for
example in a GLUT application. To keep this separation, an external
'mirror' hierarchy is constructed around the tree. This hierarchy of
\textsf{VertexBufferDist} nodes is responsible for cloning the model
data on the remote render clients.

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/modelDist.pdf}
  {\caption{\label{fModelDist}Scene Data in eqPly}}
\end{wrapfigure}
The identifier of the model's root object of this distributed hierarchy
is passed as part of the \textsf{InitData} for the default model, or as
part of the \textsf{View} for each logical view. It is used on the
render clients to map the model when it is needed for
rendering. \fig{fModelDist} shows the UML hierarchy of the model and
distribution classes.

Each \textsf{VertexBufferDist} object corresponds to one node of the
model's data tree. It is serializing the data for this
node. Furthermore, it mirrors the kD-tree by having a
\textsf{VertexBufferDist} child for each child of its corresponding tree
node. During serialization, the identifier of these children is sent to
the remote nodes, which reconstruct the mirror distribution hierarchy
and model data tree based on this data.

@@

\begin{wrapfigure}{r}{.618\textwidth}
%\begin{figure}[ht!]\center
  \vspace{-2ex}
  \includegraphics[width=.618\textwidth]{images/objects.pdf}
  {\caption{\label{fObjects}Scene Graph Distribution}}
\end{wrapfigure}
Applications distributing a dynamic scene graph use the frame data
instead of the init data as the entry point to their scene graph data
structure. \fig{fObjects} shows one possible implementation, where the
identifier and version of the scene graph root are transported using the
frame data. The scene graph root then serializes and de-serializes his
immediate children by transferring their identifier and current version,
similar to the static distribution done by \textsf{eqPly}.

@@

The config initialization happens in parallel, that is, all config
initialization tasks are transmitted by the server at once and their
completion is synchronized afterwards.

%\begin{figure}[ht!]\center
\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/configInit.pdf}
  {\caption{\label{fConfigInit}Config Initialization Sequence}}
\end{wrapfigure}
The tasks are executed by the node and pipe threads in parallel. The
parent's initialization methods are always executed before any child
initialization method. This parallelization is necessary to allow a
speedy startup of the configuration on large-scale graphics clusters. On
the other hand, it means that initialization functions are called even
if the parent's initialization has failed. \fig{fConfigInit} shows a
sequence diagram of the config initialization.

@@

\subsubsection{Model Handling}

Model data is handled by the config. Models in eqPly are static,
and therefore the render clients only need to map one instance of the
model per node.

Multiple models can be loaded in \textsf{eqPly}. A configuration has a
default model, stored in \textsf{InitData}, and one model per view,
stored and distributed using \textsf{View}. The
loaded models are evenly distributed over the available views of the
configuration, as shown in \fig{fLayout}.

The channel acquires the model during rendering from the config, using
the model identifier from its current view, or from the frame data if
no view is configured.

The config maintains the mapped models, and lazily maps new models,
which are registered by the application process:

{\footnotesize\begin{lstlisting}
const Model* Config::getModel( const uint32_t modelID )
{
    if( modelID == EQ_ID_INVALID ) // no model loaded by application
        return 0;

    // Accessed concurrently from pipe threads
    eq::base::ScopedMutex< eq::base::SpinLock > _mutex( _modelLock );

    const size_t nModels = _models.size();
    EQASSERT( _modelDist.size() == nModels );

    for( size_t i = 0; i < nModels; ++i )
    {
        const ModelDist* dist = _modelDist[ i ];
        if( dist->getID() == modelID )
            return _models[ i ];
    }

    _modelDist.push_back( new ModelDist );
    Model* model = _modelDist.back()->mapModel( this, modelID );
    EQASSERT( model );
    _models.push_back( model );

    return model;
}
\end{lstlisting}}

\subsubsection{Layout and View Handling}

For layout and model selection, \textsf{eqPly} maintains an active
view. The identifier of the active view is stored in the frame data,
which is used by the render client to highlight it using a different
background color. The active view can be selected by clicking into a
view, or by cycling through all the views using a keyboard shortcut:

{\footnotesize\begin{lstlisting}
bool Config::handleEvent( const eq::ConfigEvent* event )
{
    switch( event->data.type )
    {
        ...
        case eq::Event::POINTER_BUTTON_PRESS:
            _frameData.setCurrentViewID( event->data.context.view.id );
            return true;
        ...
    }
}

bool Config::_handleKeyEvent( const eq::KeyEvent& event )
{
    switch( event.key )
    {
        ...
        case 'v':
        case 'V':
        {
            NextViewFinder finder( _frameData.getCurrentViewID( ));
            accept( finder );
            const eq::View* view = finder.getResult();
            if( view )
                _frameData.setCurrentViewID( view->getID( ));
            else
                _frameData.setCurrentViewID( EQ_ID_INVALID );
            return true;
        }
        ...
    }
}
\end{lstlisting}}

The model of the active view can be changed using a keyboard
shortcut. The model is view-specific, and therefore the model identifier
for each view is stored on the view, which is used to retrieve the model
on the render clients. View-specific data is not limited to a
model. Applications can choose to make any application-specific data
view-specific, e.g., cameras, rendering modes or annotations. A view is
a generic concept for an application-specific view on data,
\textsf{eqPly} is using different models to illustrate the concept:

{\footnotesize\begin{lstlisting}
bool Config::_handleKeyEvent( const eq::KeyEvent& event )
{
    switch( event.key )
    {
        ...
        case 'm':
        case 'M':
        {
            if( _modelDist.empty( )) // no models
                return true;

            // current model
            const uint32_t viewID = _frameData.getCurrentViewID();
            View* view = static_cast< View* >( findView( viewID ));
            const uint32_t currentID = view ? 
                view->getModelID() : _frameData.getModelID();

            // next model
            ModelDistVector::const_iterator i;
            for( i = _modelDist.begin(); i != _modelDist.end(); ++i )
            {
                if( (*i)->getID() != currentID )
                    continue;

                ++i;
                break;
            }
            if( i == _modelDist.end( ))
                i = _modelDist.begin(); // wrap around

            // set identifier on view or frame data (default model)
            const uint32_t modelID = (*i)->getID();
            if( view )
                view->setModelID( modelID );
            else
                _frameData.setModelID( modelID );
            return true;
        }
        ...
    }
}
\end{lstlisting}}

The layout of the canvas with the active view can also be dynamically
switched using a keyboard shortcut. The first canvas using the layout is
found, and then the next layout of the configuration is set on this
canvas.

Switching a layout causes the initialization and de-initialization task
methods to be called on the involved channels, and potentially windows,
pipes and nodes. This operation might fail, which will cause the config
to stop running.

Layout switching is typically used to change the presentation of views
at runtime. The source code omitted for brevity.

@@

For each active render client, one \textsf{eq::Node} instance is
created on the appropriate machine. Nodes are only instantiated on their
render client processes, i.e., each process will only have one
instance of the \textsf{eq::Node} class. The application process might
also have a node class, which is handled in exactly the same way as the
render client nodes. The application and render clients might use a
different node factory, instantiating a different type of \textsf{eq::Node}.

@@

The actual mapping of the static data is done by the config. The config
retrieves the distributed \textsf{InitData}. The object is directly
unmapped since it is static, and the data has been retrieved:

{\footnotesize\begin{lstlisting}
bool Config::mapData( const uint32_t initDataID )
{
    if( _initData.getID() == EQ_ID_INVALID )
    {
        EQCHECK( mapObject( &_initData, initDataID ));
        unmapObject( &_initData ); // data was retrieved, unmap immediately
    }
    else  // appNode, _initData is registered already
        EQASSERT( _initData.getID() == initDataID );
}
\end{lstlisting}}

@@

The object manager is, strictly speaking, not a part of the window. It
is mentioned here since the \textsf{eqPly} window uses an object manager.

@@

\textsf{eqPly} uses the window's object manager in the rendering code to
obtain the OpenGL objects for a given data item. The address of the data
item to be rendered is used as the key.

For the currently supported types of OpenGL objects please refer to the
API documentation on the Equalizer website. For each object, the
following functions are available:

\begin{description}
\item[supportsObjects()] returns true if the usage for this particular
  type of objects is supported. For objects available in OpenGL 1.1 or
  earlier, this function is not implemented.
\item[getObject( key )] returns the object associated with the given
  key, or FAILED.
\item[newObject( key )] allocates a new object for the given
  key. Returns FAILED if the object already exists or if the allocation
  failed.
\item[obtainObject( key )] convenience function which gets or obtains
  the object associated with the given key. Returns FAILED only if the
  object allocation failed.
\item[deleteObject( key )] deletes the object.
\end{description}

@@

\begin{wrapfigure}{r}{.382\textwidth}
  \vspace{-8ex}\includegraphics[width=.382\textwidth]{images/DB.png}
  {\caption{\label{fDBDest}Destination view of a DB compound}}
\end{wrapfigure}
After the basic view setup, a directional light is configured, and the
model is positioned using the camera parameters from the frame data. The
camera parameters are transported using the the frame data to ensure
that all channels render a given frame using the same position.

@@

    Config*          config = static_cast< Config* >( getConfig( ));
    const Model*     model  = getModel();
    const eq::Range& range  = getRange();

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/render.pdf}
  {\caption{\label{fRender}Main Render Loop}}
\end{wrapfigure}
The rendering loop maintains a list of candidates to render, which
initially contains the root node. Each candidate of this list is tested
for full visibility against the frustum and range, and rendered if
visible. It is dropped if it is fully invisible or fully out of
range. If it is partially visible or partially in range, the children of
the node are added to the candidate list. \fig{fRender} shows a flow
chart of the rendering algorithm, which performs efficient view frustum
and range culling.

The actual rendering uses display lists or vertex buffer objects. These
OpenGL objects are allocated using the object manager. The rendering is
done by the leaf nodes, which are small enough to store the vertex
indices in a \textsf{short} value for optimal performance with VBO's.
The leaf nodes reuse the objects stored in the object manager, or create
and set up new objects if it was not yet set up. Since one object
manager is used per thread (pipe), this allows a thread-safe sharing of
the compiled display lists or VBO's across all windows of a pipe.

@@

The purpose of the \textsf{OSWindow} processing method,
\textsf{processEvent}, is to perform window-system specific event
processing. For example, \textsf{AGLWindow::processEvent} calls
\textsf{aglUpdateContext} whenever a resize event is received. For
further, generic processing, the \textsf{Event} is passed on to
\textsf{Window::processEvent}. This \textsf{Event} no longer contains
the native event.

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/eventFilter.pdf}
  {\caption{\label{fEventProcessing}Event Processing}}
\end{wrapfigure}
\textsf{Window::processEvent} is responsible for either handling
the event locally, or for translating it into a generic
\textsf{ConfigEvent}. The config events are send to the application
thread using \textsf{Config::sendEvent}.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/eventUML.pdf}
  {\caption{\label{fEventUML}UML Diagram for Event Handling Classes}}
\end{wrapfigure}
The concrete implementation of these window interfaces is responsible
for setting up and de-initializing event handling.

@@

The glX event handler iterates over all windows of the registered pipe
to find the originating \textsf{GLXWindowIF}, which is then used to
process the event in a fashion similar to AGL. The
\textsf{GLXWindowEvent} passes the \textsf{XEvent} to
\textsf{GLXWindowIF::\-process\-Event}.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/threads.pdf}
  {\caption{\label{fThreads}Threads within one node process}}
\end{wrapfigure}
During this initialization, Equalizer creates and manages two threads
for communication, the receiver thread and the command thread. Normally
no application code is executed from these two threads.

@@

The receiver and command thread are terminated when the application
stops network communications using \textsf{Client::exitLocal}.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/frameSync.pdf}
  {\caption{\label{fFrameSync}Per-node frame synchronization}}
\end{wrapfigure}
The application has extended control over the task synchronization
during a frame. Upon \textsf{Config::startFrame}, Equalizer invokes the
\textsf{frameStart} task methods of the various entities. The entities
unlock all their children by calling \textsf{startFrame}, e.g.,
\textsf{Node::frameStart} has to call \textsf{Node::startFrame} in order
to unlock the pipe threads. Note that certain \textsf{startFrame} calls,
e.g., \textsf{Window::startFrame}, are currently empty since the
synchronization is implicit due to the sequential execution within the
thread.

@@

The local synchronization consists of \textsf{releaseFrameLocal} to
unlock the local frame, and of \textsf{waitFrameLocal} to wait for the
unlock. For the default synchronization model \textsf{sync\_draw},
Equalizer uses the task method \textsf{frameDrawFinish} which is called
on each resource after the last \textsf{Channel::frameDraw} invocation
for this frame. Consequently, \textsf{Pipe::frameDraw\-Fi\-nish} calls
\textsf{Pipe::\-release\-Frame\-Lo\-cal} to signal that it is done
drawing the current frame, and \textsf{Node::frameDrawFinish} calls
\textsf{Pipe::waitFrameLocal} for each of its pipes to block the node
thread until the current frame has been drawn.

@@

\begin{enumerate}
\item\textsf{initWGLAffinityDC} is used to set up an affinity device
  context, which might be needed for window creation. The WGL window
  tracks potentially two device context handles, one for OpenGL context
  creation (the affinity DC), and one for \textsf{swapBuffers} (the
  window's DC).
\item\textsf{chooseWGLPixelFormat} chooses a pixel format based on the
  window attributes. If no device context is given, it uses the system
  device context. The chosen pixel format is set on the passed device
  context.
\item\textsf{configInitWGLDrawable} creates the drawable. The device
  context passed to \textsf{configInitWGLDrawable} is used to query the
  pixel format and is used as the device context for creating a
  PBuffer. If no device context is given, the display device context is
  used. On success, it sets the window handle. Setting a window handle
  also sets the window's device context.
\item\textsf{createWGLContext} creates an OpenGL rendering context using
  the given device context. If no device context is given, the window's
  device context is used. This function does not set the window's OpenGL
  context.
\end{enumerate}

@@

    if( !joinNVSwapBarrier( ))
    {
        _window->setErrorMessage( "Joining NV_swap_group failed" );
        return false;
    }
